Water, High-Altitude Condensates, and \added{Possible} Methane Depletion in the Atmosphere of the Warm Super-Neptune WASP-107\lowercase{b}

### 2018-02-25 ApJL_submission/ms.tex
clouds

\subsection{Spectroscopic Light Curve Fits}
We binned the spectrum into 20 spectrophotometric channels from 1.12 to 1.63 $\mu$m, shown in Figure\,\ref{fig:lc}. We fit the light curves with the \texttt{divide-white} technique, which assumes that the light curve systematics have the same morphology at all wavelengths \citep{stevenson14c, kreidberg14a}. For this method, the transit model $T_\lambda(t)$ is multiplied by the systematics vector from the white light curve fit ($F_\mathrm{white}/T_\mathrm{white}$), and rescaled by a factor $C_\lambda + V_\lambda t$.  One advantage of this approach is that it removes the star spot crossing feature, enabling us to use orbit three with no additional correction. The amplitude of the feature has no detectable wavelength dependence at the level of precision of our data.  As for the white light curve, we fixed some of the transit parameters on the best fit values from \cite{dai17}. We did not put put priors on the transit parameters because varying them simply shifts the spectrum up or down by a constant value. Since the planet's 10-bar radius is a free parameter in our analysis of the spectrum, the absolute transit depth does not affect the retrieved atmospheric properties. In addition to fixing the transit parameters, we also fixed the limb darkening on the PHOENIX model predictions (coefficients listed in Table\,\ref{tab:tspec}.  The final spectroscopic light curve fits had just three free parameters per channel: $C_\lambda$, $V_\lambda$, and $r_p/r_s$.  We refer to this fitting approach as ``Method A".

@@

We explored several alternative choices for the spectroscopic light curve fits, but found that none of them made a significant difference in the transmission spectrum. In one test, we fit the spectroscopic light curves with the same analytic model we used for the white light curve. \added{This fit (which we label ``Method B") has additional free parameters, so the transit depth uncertainties increase by 50\%, but the best fit transit depths are consistent to within $0.5\sigma$ on average over the bandpass.} We also tested a white light curve systematics vector $S_\mathrm{white}$ from a fit that included the star-spot crossing orbit. The results from these tests are nearly identical to our nominal transmission spectrum, except with a small constant offset due to the uncorrected star-spot crossing feature. This offset does not affect our final analysis because it is marginalized in the atmospheric retrieval with the 10-bar radius parameter (see \S\,\ref{sec:retrieval}). 
	\added{We also repeated the calculations from Method A, but fit} for a linear limb darkening parameter rather than fixing the limb darkening on the PHOENIX model values. \added{This approach, which we dub ``Method C", also yields consistent results with Method A (the relative transit depths agree to better than 0.2 sigma on average). The transit depths and fitted limb darkening coefficients are provide in Table\,\ref{tab:tspec}.} The fitted limb darkening coefficients are consistent with the model predictions, so we opted to fix the coefficients in our final analysis because it improves the precision on the transit depths.  \added{We calculated model limb darkening coefficients for WASP-107 over the $1\,\sigma$ range of stellar effective temperatures, surface gravities, and metallicities published in the discovery paper ($T_\mathrm{eff} = 4430\pm130$ K, $\log g = 4.5 \pm 0.1$, and $[\mathrm{Fe/H}] = 0.02 \pm 0.1i$). The uncertainty in $T_\mathrm{eff}$ has the largest effect on the estimated limb darkening coefficients. We recalculated the transmission spectrum with limb darkening models that varied $T_\mathrm{eff}$ from 4310 to 4550 K. Over this temperature range, the change in limb darkening shifts the average transit depth by less than 20 ppm, which is below the level of noise in the data.  The relative change in transit depth is smaller still ($< 1$ ppm).}

In addition to these tests, we also performed an independent data reduction and fit using K. Stevenson's pipeline and found consistent results (well within $1\,\sigma$).

\added{\subsubsection{Impact of Unocculted Star Spots}
	Unocculted star spots may influence the measured transmission spectrum \citep{mccullough14, zellem17}. To estimate the impact of unocculted spots on our data,  we first estimated the star spot properties based on the observed photometric variability in the Kepler band pass, which is 0.3\% \citep{dai17,mocnik17}. This variability can be reproduced with a single spot with covering fraction of 1\% and temperature 300 K lower than the stellar photosphere (4130 vs 4430 K).  We calculated the effect of unocculted spots with these properties using equation (1) from \cite{mccullough14}, and predict that they will introduce a positive slope in the transmission spectrum of 8 ppm over the WFC3 bandpass, well below our measurement uncertainties. We note that cool stars may have persistent spot coverage that does not produce any photometric variability \citep{rackham18}. However, even if there is persistent spot coverage of 5\%, the spots produce a slope in the WFC3 bandpass of 50 ppm, which is still below our uncertainties.}

@@

We explored a few retrieval scenarios with additional complexity, including the addition of cloud patchiness \citep{line16}, a quench pressure for nitrogen and carbon species \citep[e.g.][]{morley17}, and a power law haze opacity.  We also varied the assumed planet mass by $3\,\sigma$. These models did not significantly improve the fit quality, and the main retrieval results were unchanged.  We also explored cases without the $30\times$ solar metallicity prior upper limit. Without this prior, \added{the retrieved metallicity is $[M/H] = 1.35^{+0.68}_{-1.48}$, log10(C/O) is $-1.57+{0.54}_{-0.31}$, and cloud top pressure $log(Pc) = -3.63^{+0.81}_{-0.48}.$ These results show that the inferred C/O ratio and cloud-top pressure are unaffected by the prior on atmospheric metallicity.  The metallicity is consistent with the interior structure modeling (which puts a $3\,\sigma$ upper limit on metallicity of $[M/H] = 1.5$). The transmission spectrum does not constrain the atmospheric metallicity more precisely because there is a degeneracy between metallicity and the planet's 10-bar radius \citep[e.g]{griffith14,heng17}.}

@@

\subsection{Condensate Properties}
In addition to the atmospheric retrieval, we also performed forward modeling of physically motivated, self-consistent clouds and hazes using the methods described in \cite{fortney08, morley15}.  We model clouds that form in cool atmospheres (Na$_2$S, KCl, ZnS; see \citealt{morley12}), for a range of metallicites from $1-50\times$ solar. We vary the cloud sedimentation efficiency from 3 to 0.1 (normal to highly lofted, small particulate clouds). \added{FIXME}.

None of the models produce sufficiently low amplitude spectral features to match the observed spectrum.  This result suggests that if clouds are muting WASP-107b's transmission spectrum, the mechanism for cloud particle formation and vertical lofting is very efficient and not captured by the modeling used here. Futher studies of cloud formation in very low gravity environments are required.

We also model an \emph{ad hoc} photochemical ``soot" layer near the top of the atmosphere. We predict the abundance of hydrocarbon haze precursors from previously published photochemical models for GJ 436b \citep{line11, morley17}. With the most efficient haze production (100\% of precursors form haze) and particle sizes around $0.03-0.1$ microns, the amplitude of the model water feature matches that of the observations.

These photochemical hazes are assumed to form from hydrocarbons generated by methane photolysis; however, we found in the previous section that methane is depleted.  Furthermore, even if the precursors were present, it is unclear whether the 100\% haze production efficiency is realistic.  However, organic hydrocarbons are not the only possible photolytic haze in cool H2-rich atmospheres, as sulfur chemistry may create additional haze material \citep{zahnle09, gao17}. Further modeling and laboratory work of the hazes that form in these conditions is needed for a satisfactory explanation of the data.

@@

\begin{itemize}
\item{\emph{The upper limit on atmospheric metallicity from interior structure modeling is $30\times$ solar.} This limit is at the edge of consistency with the Solar System metal enrichment trend, which predicts a metallicity of $30\times$ solar for WASP-107b \citep{kreidberg14b}.  Compared to results for other exoplanets of similar mass such as GJ 436b, which has a high metallicity ($>100\times$ solar) and HAT-P-26b, which is metal-poor compared to the Solar System trend, WASP-107b adds to the evidence that exoplanets exhibit a greater diversity of compositions than is present in the Solar System \citep{morley17, wakeford17}.}
\item{\emph{The methane abundance is \added{likely} depleted relative to expectations for a solar abundance pattern, whereas water is consistent with solar composition.} This may be due to an instrinsically low carbon-to-oxygen ratio, which could arise from accretion of water-rich planetesimals \citep{mordasini16, espinoza17}.  Another possibility is that the planet has a hot interior effective temperature ($\sim500$ K), and abundances are quenched at pressure levels where CO is stable \citep[as observed in some directly imaged planets;][]{skemer14, zahnle14}. Such a high internal temperature could be due to latent heat of formation if the planet is very young, and/or tidal heating \citep{fortney08, morley17}. Further observations of the transmission spectrum over a broader wavelength range will refine the C/O estimate and help differentiate between these two scenarios.} 
\item{\emph{Optically thick condensates are present at high altitudes} ($0.1 - 3$ mbar). The amplitude of the water absorption feature in the transmission spectrum is less than a third that expected for a clear atmosphere. Existing models of physically-motivated cloud and haze formation do not satisfactorily reproduce the data. All the cloud models we consider produce condensation too deep in the atmosphere. Efficient hydrocarbon haze production at high altitudes can match the data, but these require the presence of methane to serve as a haze precursor.  Other precursors such as H$_2$S are a possibility and would be worth exploring in future analyses. Put in context with other systems, the muted water feature for WASP-107b agrees well with the trend in feature amplitude with temperature noted in \cite{crossfield17}, indicating that condensates may be common in the atmospheres of the coolest planets.}
\end{itemize}

>>>>

\subsection{Spectroscopic Light Curve Fits}
We binned the spectrum into 20 spectrophotometric channels from 1.12 to 1.63 $\mu$m, shown in Figure\,\ref{fig:lc}. We fit the light curves with the \texttt{divide-white} technique, which assumes that the light curve systematics have the same morphology at all wavelengths \citep{stevenson14c, kreidberg14a}. For this method, the transit model $T_\lambda(t)$ is multiplied by the systematics vector from the white light curve fit ($F_\mathrm{white}/T_\mathrm{white}$), and rescaled by a factor $C_\lambda + V_\lambda t$.  One advantage of this approach is that it removes the star spot crossing feature, enabling us to use orbit three with no additional correction. The amplitude of the feature has no detectable wavelength dependence at the level of precision of our data.  As for the white light curve, we fixed some of the transit parameters on the best fit values from \cite{dai17}. We did not put put priors on the transit parameters because varying them simply shifts the spectrum up or down by a constant value. Since the planet's 10-bar radius is a free parameter in our analysis of the spectrum, the absolute transit depth does not affect the retrieved atmospheric properties. In addition to fixing the transit parameters, we also fixed the limb darkening on the PHOENIX model predictions \added{for a star with $T_\mathrm{eff} = 4300 K$, $\log g = 4.5$, and $[\mathrm{Fe/H}] = 0.$} (coefficients listed in Table\,\ref{tab:tspec}.  The final spectroscopic light curve fits had just three free parameters per channel: $C_\lambda$, $V_\lambda$, and $r_p/r_s$.  We refer to this fitting approach as ``Method A".

@@

We explored several alternative choices for the spectroscopic light curve fits, but found that none of them made a significant difference in the transmission spectrum. In one test, we fit the spectroscopic light curves with the same analytic model we used for the white light curve. \added{This fit (which we label ``Method B") has additional free parameters, so the transit depth uncertainties increase by 50\%, but the best fit transit depths are consistent within $0.5\sigma$ on average with method A over the bandpass.} We also tested a white light curve systematics vector $S_\mathrm{white}$ from a fit that included the star-spot crossing orbit. The results from these tests are nearly identical to our nominal transmission spectrum, except with a small constant offset due to the uncorrected star-spot crossing feature. This offset does not affect our final analysis because it is marginalized in the atmospheric retrieval with the 10-bar radius parameter (see \S\,\ref{sec:retrieval}). 
	\added{We also repeated the calculations from Method A, but fit} for a linear limb darkening parameter rather than fixing the limb darkening on the PHOENIX model values. \added{This approach, which we dub ``Method C", also yields consistent results with Method A (the relative transit depths agree to better than 0.2 sigma on average). The transit depths and fitted limb darkening coefficients are provide in Table\,\ref{tab:tspec}.} The fitted limb darkening coefficients are consistent with the model predictions, so we opted to fix the coefficients in our final analysis because it improves the precision on the transit depths.  
	\added{To test that uncertainties on the stellar parameters do not significantly bias the limb darkening coefficients, we calculated PHOENIX model limb coefficients for WASP-107 over the $1\,\sigma$ range of stellar effective temperatures, surface gravities, and metallicities published in the discovery paper ($T_\mathrm{eff} = 4430\pm130$ K, $\log g = 4.5 \pm 0.1$, and $[\mathrm{Fe/H}] = 0.02 \pm 0.1$). The uncertainty in $T_\mathrm{eff}$ has the largest effect on the estimated limb darkening coefficients. We recalculated the transmission spectrum with limb darkening models that varied $T_\mathrm{eff}$ from 4310 to 4550 K. Over this temperature range, the change in limb darkening shifts the average transit depth by less than 20 ppm, which is below the level of noise in the data.  The relative change in transit depth is smaller still ($< 1$ ppm).}

In addition to these tests, we also performed an independent data reduction and fit using K. Stevenson's pipeline and again found consistent results (well within $1\,\sigma$).

\added{\subsubsection{Impact of Unocculted Star Spots}
	Unocculted star spots may influence the measured transmission spectrum \citep{mccullough14, zellem17}. To estimate the impact of unocculted spots on our data,  we first estimated the star spot properties based on the observed photometric variability in the Kepler band pass, which is 0.3\% \citep{dai17,mocnik17}. This variability can be reproduced with a single spot with covering fraction of 1\% and temperature 300 K lower than the stellar photosphere.  We calculated the effect of unocculted spots with these properties using equation (1) from \cite{mccullough14}, and predict that they will introduce a positive slope in the transmission spectrum of 8 ppm over the WFC3 bandpass, well below our measurement uncertainties. We note that cool stars may have persistent spot coverage that does not produce any photometric variability \citep{rackham18}. However, even if there is persistent spot coverage of 5\%, the spots produce a slope in the WFC3 bandpass of 50 ppm, which is still below our uncertainties.}

@@

We explored a few retrieval scenarios with additional complexity, including the addition of cloud patchiness \citep{line16}, a quench pressure for nitrogen and carbon species \citep[e.g.][]{morley17}, and a power law haze opacity.  We also varied the assumed planet mass by $3\,\sigma$. These models did not significantly improve the fit quality, and the main retrieval results were unchanged.  We also explored cases without the $30\times$ solar metallicity prior upper limit. Without this prior, \added{the retrieved metallicity is $[M/H] = 1.35^{+0.68}_{-1.48}$, log10(C/O) is $-1.57^{+0.54}_{-0.31}$, and cloud top pressure $\log P_\mathrm{c} = -3.63^{+0.81}_{-0.48}.$ These results show that the inferred C/O ratio and cloud-top pressure are unaffected by the prior on atmospheric metallicity.  The metallicity is consistent with the interior structure modeling (which puts a $3\,\sigma$ upper limit on metallicity of $[\mathrm{M/H}] = 1.5$). The transmission spectrum does not constrain the atmospheric metallicity more precisely because there is a degeneracy between metallicity and the planet's 10-bar radius \citep[e.g]{griffith14,heng17} with the constraint primarily driven by the H$_2$O-H$_2$ CIA continuum ratio, molecular weight \citep{line16}, and pressure broadening \citep{dewit14}.}

@@

\subsection{Condensate Properties}
In addition to the atmospheric retrieval, we also performed forward modeling of physically motivated, self-consistent clouds and hazes using the methods described in \cite{fortney08, morley15}.  We modeled clouds that form in cool atmospheres (Na$_2$S, KCl, ZnS; see \citealt{morley12}), for a solar composition atmosphere. We varied the cloud sedimentation efficiency from 3 to 0.3 (normal to highly lofted, small particulate clouds).  \added{Only the model with the most lofted clouds ($f_\mathrm{sed} = 0.3$) produces sufficiently small amplitude features to match the observed spectrum.  This result is in agreement with findings for other small planets, such as the low sedimentation efficiency inferred for GJ 1214b \citep[$f_\mathrm{sed} < 0.1$][]{morley15}.}  
%cite Peter Gao's paper?

%This result suggests that if clouds are muting WASP-107b's transmission spectrum, the mechanism for cloud particle formation and vertical lofting is very efficient and not captured by the modeling used here. Futher studies of cloud formation in very low gravity environments are required.

We also modeled an \emph{ad hoc} photochemical ``soot" layer near the top of the atmosphere. We predicted the abundance of hydrocarbon haze precursors from previously published photochemical models for GJ 436b \citep{line11, morley17}. \added{We ran haze models for two different particle sizes ($0.03$ and $0.3\,\mu$m), for haze production efficiencies $f_\mathrm{haze}$ of 3, 10, and 30\%. We found that the larger particles ($0.3\,\mu$m) with moderate haze production efficiency (3\%) match the shape and amplitude of the spectral features well. More efficient haze production produces features smaller than we measure, and smaller particles sizes produce a slope (an increase in transit depth towards bluer wavelengths) that is not seen.}

%These photochemical hazes are assumed to form from hydrocarbons generated by methane photolysis; however, we found in the previous section that methane is depleted.  Furthermore, even if the precursors were present, it is unclear whether the 100\% haze production efficiency is realistic.  However, organic hydrocarbons are not the only possible photolytic haze in cool H2-rich atmospheres, as sulfur chemistry may create additional haze material \citep{zahnle09, gao17}. Further modeling and laboratory work of the hazes that form in these conditions is needed for a satisfactory explanation of the data.

@@

\begin{itemize}
\item{\emph{The upper limit on atmospheric metallicity from interior structure modeling is $30\times$ solar.} This limit is at the edge of consistency with the Solar System metal enrichment trend, which predicts a metallicity of $30\times$ solar for WASP-107b \citep{kreidberg14b}.  Compared to results for other exoplanets of similar mass such as GJ 436b, which has a high metallicity ($>100\times$ solar) and HAT-P-26b, which is metal-poor compared to the Solar System trend, WASP-107b adds to the evidence that exoplanets exhibit a greater diversity of compositions than is present in the Solar System \citep{morley17, wakeford17}.}
\item{\emph{The methane abundance is \added{likely} depleted relative to expectations for a solar abundance pattern, whereas water is consistent with solar composition.} This may be due to an instrinsically low carbon-to-oxygen ratio, which could arise from accretion of water-rich planetesimals \citep{mordasini16, espinoza17}.  Another possibility is that the planet has a hot interior effective temperature ($\sim500$ K), and abundances are quenched at pressure levels where CO is stable \citep[as observed in some directly imaged planets;][]{skemer14, zahnle14}. Such a high internal temperature could be due to latent heat of formation if the planet is very young, and/or tidal heating \citep{fortney08, morley17}. Further observations of the transmission spectrum over a broader wavelength range will refine the C/O estimate and help differentiate between these two scenarios.} 
\item{\emph{Optically thick condensates are present at high altitudes} ($0.1 - 3$ mbar). The amplitude of the water absorption feature in the transmission spectrum is less than a third that expected for a clear atmosphere.  \added{We find that physically-motivated cloud and haze formation models can satisfactorily reproduce the observed feature amplitude. Either lofted clouds with low sedimentation efficiency ($f_\mathrm{sed} = 0.3$) or $0.3\,\mu$m haze particles with moderate haze production efficiency (3\%) match the spectrum.} Put in context with other systems, the muted water feature for WASP-107b agrees well with the trend in feature amplitude with temperature noted in \cite{crossfield17}, indicating that condensates may be common in the atmospheres of the coolest planets.}
\end{itemize}
